---
title: "class19"
author: 'Groot (PID: A15485151)'
date: "11/30/2021"
output: pdf_document
---


#Proportion of G/G in a population
```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

```{r}
table(mxl$Genotype..forward.strand.)

```
```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) * 100

```


##Section 4: Population Scale Analysis
> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
table(expr$geno)
```
**sample size**:
  A/A 108
  A/G 233
  G/G 121
  

```{r}
library(ggplot2)
```
```{r}
box <- ggplot(expr)+ aes(geno, exp, fill= geno) + geom_boxplot(notch = TRUE)
box
```

**median expression level**:
  A/A 32
  A/G 25
  G/G 20
  

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
```
```{r}
ggplot(expr)+ aes(geno, exp, fill= geno) + geom_boxplot(notch = TRUE)

```
The relative gene expression is reduced from A/A to G/G so we could infer that the G/G genotype variation or the G/G SNP does affect the expression level of ORMDL3.



