---
title: "Class 16 DESeq2 mini-project"
author: 'Groot (PID: A15485151)'
date: "11/16/2021"
output: pdf_document
---

load DESeq2
```{r}
library(DESeq2)
```

load data files
```{r}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"

# Import metadata and take a peek
colData = read.csv(metaFile, row.names = 1)
head(colData)
```
Import count data
```{r}
countData = read.csv(countFile, row.names=1)
head(countData)
```

> Q. Complete the code below to remove the troublesome first column from countData

```{r}
# Note we need to remove the odd first $length col
countData <- as.matrix(countData[,-1])
head(countData)
```

> Q. Complete the code below to filter countData to exclude genes (i.e. rows) where we have 0 read count across all samples (i.e. columns).


```{r}
# Filter count data where you have 0 read count across all samples.
to.rm <- which(rowMeans(countData) == 0, arr.ind = TRUE)
countData2 = countData[-to.rm, ]
head(countData2)
```

##Running DESeq2

```{r}
dds = DESeqDataSetFromMatrix(countData=countData2,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)
```

```{r}
dds
```

Get results for the HoxA1 knockdown versus control siRNA (remember that these were labeled as "hoxa1_kd" and "control_sirna" in our original colData metaFile input to DESeq

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))
```

> Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
summary(res)
```

reset p-value to 0.05
```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

##Volcano plot

```{r}
plot( res$log2FoldChange, -log(res$padj) )
```

> Q. Improve this plot by completing the below code, which adds color and axis labels

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

##Adding gene annotation
> Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.


```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)
```

```{r}
res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

> Q. Finally for this section let's reorder these results by adjusted p-value and save them to a CSV file in your current project directory.

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```

##Section 2. Pathway Analysis

#KEGG pathways
```{r}
library(pathview)
```

```{r}
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

# Focus on signaling and metabolic pathways only
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]

# Examine the first 3 pathways
head(kegg.sets.hs, 3)
```

The main gage() function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

run the gage pathway analysis.
```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

look at the object returned from gage().
```{r}
attributes(keggres)
```

look at the first few down (less) pathway results:
```{r}
head(keggres$less)
```

use the pathview() function to make a pathway plot for the pathway 'Cell Cyle' (hsa04110)

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

You can play with the other input arguments to pathview() to change the display in various ways including generating a PDF graph
```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
```

Now, let's process our results a bit more to automagicaly pull out the top 5 upregulated pathways, then further process that just to get the pathway IDs needed by the pathview() function. We'll use these KEGG pathway IDs for pathview plotting below

```{r}
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids
```

Pass these IDs in keggresids to the pathview() function to draw plots for all the top 5 pathways
```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

##Section 3. Gene Ontology (GO)
```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

##Section 4. Reactome Analysis

Let's now conduct over-representation enrichment analysis and pathway-topology analysis with Reactome using the previous list of significant genes generated from our differential expression results above.

First, Using R, output the list of significant genes at the 0.05 level as a plain text file:
```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

Then, to perform pathway analysis online go to the Reactome website (https://reactome.org/PathwayBrowser/#TOOL=AT). Select “choose file” to upload your significant gene list. Then, select the parameters “Project to Humans”, then click “Analyze”.

> Q: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

Endosomal/vacuolar pathway has the most significant "Entities p-value" and this does not match the previous KEGG results. One of the factors that could cause difference between KEGG and Reactome analysis is the fact that these two methods use different databases so the possible, overlapping pathways that can be identified might be different. Another factor is the difference in the way these two methods do the math in overlapping and networking the pathways.




