---
title: "Vaccine rate mini project"
author: 'Groot (PID: A15485151)'
date: "11/23/2021"
output: 
  pdf_document:
      latex_engine: xelatex

---

Get started
```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

##Ensure the date column is useful 

We will use the **lubridate** package to make life a lot easier wehn dealing with dates and times

```{r}
#install.packages("lubridate")
library(lubridate )
```

```{r}
today()
```

Here we make our 'as_of_date' column lubridate format 

```{r}
# Specify that we are using the Year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

Now I can do useful math with dates easily:

> Q. How many days since the first entry?

```{r}
today() - vax$as_of_date[1]
```
322 days

> Q. How many days since the last entry?

```{r}
today() - vax$as_of_date[nrow(vax)]
```
7 days

> Q1. What column details the total number of people fully vaccinated? 

```{r}
colnames(vax)
```
"persons_fully_vaccinated" 

> Q2. What column details the Zip code tabulation area? 

"zip_code_tabulation_area"

> Q3. What is the earliest date in this dataset? 

```{r}
min(vax$as_of_date) 
```
"2021-01-05"

> Q4. What is the latest date in this dataset? 

```{r}
max(vax$as_of_date) 
```
"2021-11-16"

```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset?

9 numeric columns

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

8256 NA values

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

10% of persons_fully_vaccinated values are missing

> Q8. [Optional]: Why might this data be missing?

maybe non-compliance to data submission/sharing?


> Q9. How many days have passed since the last update of the dataset? 

```{r}
today() - vax$as_of_date[nrow(vax)]
```

7 days

> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

46 uniques dates


##Working with zip codes

We will use the **zipcodeR** package to help make sense of zip codes

```{r}
#install.packages("zipcodeR")
library(zipcodeR)
geocode_zip('92037')
```
Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')

```


More usefully, we can pull census data about ZIP code areas (including median household income etc.). For example:

```{r}
reverse_zipcode(c('92037', "92109") )
```


We can use this reverse_zipcode() to pull census data later on for any or all ZIP code areas we might be interested in.
```{r}
# Pull data for all ZIP codes in the dataset
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )

```


#Focus on San Diego County

We will subset with base R
```{r}
inds <- vax$county == "San Diego"

head(vax[inds,])

```
But let's use the **dplyr** package and it's **filter()*** function:

```{r}
library(dplyr)
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```
107 distinct zip codes

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd[which.max(sd$age12_plus_population),]
```
92154

What is the population in the 92037 ZIP code area?

```{r}
filter(sd, zip_code_tabulation_area == "92037")[1,]
```
36144


>Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 

```{r}
mean((filter(vax, county == "San Diego" & as_of_date == "2021-11-09"))$percent_of_population_fully_vaccinated, na.rm = TRUE)
```
67%

```{r}
sd.now <- filter(sd, as_of_date == "2021-11-09")
mean(sd.now$percent_of_population_fully_vaccinated, na.rm = TRUE)
```
67%

We can look at the 6-number summary
```{r}
sd.sum <- summary(sd.now$percent_of_population_fully_vaccinated)
sd.sum
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

```{r}
library(ggplot2)
ggplot(sd.now) +
  aes(percent_of_population_fully_vaccinated)+geom_histogram(bin=15)
```


What about 92037 - UCSD/ La Jolla
```{r}
LJ <- filter(sd.now, zip_code_tabulation_area == "92037")
LJ
```
91% fully vaccinated

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x= "Date", y="Percent Vaccinated")
```


##Time series of vaccination rate for 92037

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean(vax.36$percent_of_population_fully_vaccinated)
```
mean = 66.3%

```{r}
ggplot(ucsd) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated) +
  geom_point() + 
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x= "Date", y="Percent Vaccinated") +
  geom_hline(yintercept = 0.66, colour = "red", linetype = 2)


```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```
Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 0.3519  0.5891  0.6649  0.6630  0.7286  1.0000 
 
> Q18. Using ggplot generate a histogram of this data.'

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated)+geom_histogram(bin=15) + 
  labs(x= "Percent Vaccinated", y="Count")+ 
  xlim(c(0,1))+
  ylim(c(0,60))
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
reverse_zipcode(c('92109', "92040") )
```

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```
92040 or Lakeside is below the average value (.6630)

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
```
92109 or San Diego is above the average value (.6630)

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144

First we need to subset the full vax dataset to include onl ZIP code areas with a population as large as 92037
```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
```

How many unique zip codse have a population as large as 92037?
```{r}
length(unique(vax.36.all$zip_code_tabulation_area))
```

```{r}
ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1.00)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccinte rate across California",
       subtitle="Only areas with a population above 36k are shown") +
  geom_hline(yintercept = 0.6630, linetype=2)
```

> Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I feel more anxious now that I see that still a huge proportion of areas are below the mean vaccination rate and realize that I should be more careful of who I spend the holidays and afterwards with.

