---
title: 'Class 6: R Functions'
author: 'Groot (PID: A15485151)'
date: "10/14/2021"
output: pdf_document
---

## A play with Rmarkdown
This is some plain text. I can make things **bold**. I can also make *things italic*.

```{r}
# This is a code chunk
plot(1:10)
```

## R functions

In today's class, we are going to write a function together that grades some students' works.

Questions for today:

> **Q1**. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

Let's start with student1 and find their average score
```{r}
mean(student1)
```

But we want to drop the lowest score...
We could try the **min()** function

```{r}
min(student1)
```

The **which.min()** function looks useful:

```{r}
which.min(student1)
```

Cool this gives the position of the lowest score
How would I use this information?

```{r}
#This would be the lowest score
student1[which.min(student1)]
```
To drop this value, I can use minus
```{r}
student1[-which.min(student1)]
```
"Let's now use mean() to get the average minus the lowest score.
```{r}
mean(student1[-which.min(student1)])
```
Well this sucks! It gives NA if a student has a missing homework
```{r}
mean(student2[-which.min(student2)])
```
We need to remove the NA elements of the vector
```{r}
mean(student2[-which.min(student2)],na.rm = TRUE)
```
This is not what we want. It dropped the 80 (i.e. the lowest number and not the NA i.e. missing homework)

Let's look at student3
```{r}
student3
```
```{r}
mean(student3[-which.min(student3)],na.rm = TRUE)
```

One new idea/approach is we could replace the NA(missing homeworks) with zero.

Let's try with student2.

```{r}
student2
```
```{r}
is.na(student2)
```
This **is.na()** function returns a logical vector where TRUE elements represent where the NA values are.

```{r}
which(is.na(student2))
```
Now let's make the NA values into zeros. Simplify!!
```{r}
student.prime <- student2
student.prime
student.prime[which(is.na(student.prime))]=0
student.prime
#or
x <- student2
student.prime <- replace(x, which(is.na(x)), 0)
```

Now we need to put this all together to get the average score, dropping the lowest score where we map NA values to zero.
```{r}
y <- student.prime
mean(y[-which.min(y)])
```
```{r}
student.prime
```
```{r}
mean(c(100,90,90,90,90,97,80))
```
Looks good! Now let's try with student3.
```{r}
x <- student3
student.prime <- replace(x, which(is.na(x)), 0)
y <- student.prime
mean(y[-which.min(y)])
```

Now we can use this as the body of my function.
```{r}
grade <- function(x){
  # Make sure our scores are all numbers
  x <- as.numeric(x)
  # Map NA values to zero
student.prime <- replace(x, which(is.na(x)), 0)
  # Find the mean without the lowest value
  y <- student.prime
mean(y[-which.min(y)])
}
```

Check
```{r}
grade(student3)
```


Now read the full gradebook CSV file.
```{r}
scores <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
scores
```
Use for one student
```{r}
grade(scores[10,])
```

Check if NA stays the same.
```{r}
as.numeric(c(1,2,NA,4,5))
```
GOOD!Now grade all students by using the **apply()** function
```{r}
apply(scores,1,grade)
```

> Q2. Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook? [3pts]

```{r}
which.max(apply(scores,1,grade))
```
GOOD JOB **Student18**!

>Q3. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained the lowest scores overall? [2pts]

```{r}
apply(scores,2,mean, na.rm=TRUE)
```
```{r}
which.min(apply(scores,2,mean, na.rm=TRUE))
```
**HW 3** was the toughest homework!

>Q4. Optional Extension: From your analysis of the gradebook, which homework was most predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
# Average grade score vector
avg.grade <- apply(scores,1,grade)
# Homework score vector
hw.score <- apply(scores,2,grade)

```


We want to optionally turn the dropping of lowest score off for grading
```{r}
grade2 <- function(x, drop.lowest=TRUE){
  # Make sure our scores are all numbers
  x <- as.numeric(x)
  # Map NA values to zero
student.prime <- replace(x, which(is.na(x)), 0)
  # Find the mean without the lowest value
  y <- student.prime
if(drop.lowest){mean(y[-which.min(y)])}
  else{mean(y)}
}
```

check if the option works!
```{r}
student1 <- c(100,100,100,90)
grade2(student1)
grade2(student1,drop.lowest = FALSE)
```





